\bf Looking at Non-Linear Dimension Reductions as Models in the Data Space

### 2024-09-26 paper.tex
Start to replace the methodology with a new example

With our goal being to make a representation of this \gD{} layout that
can be lifted into high-dimensional space, the layout needs to be
augmented to include neighbour information. A simple approach would be
to triangulate the points and add edges. A more stable approach is to
first bin the data, reducing it from \(n\) to \(m\leq n\) observations,
and connect the bin centroids. We recommend using a hexagon grid because
it better reflects the data distribution and has less artifacts than a
rectangular grid. This process serves to reduce some noisiness in the
resulting surface shown in \pD{}. The steps in this process are shown in
Figure~\ref{fig-NLDR-scurve}, and documented below.

@@

\includegraphics[width=1\textwidth,height=\textheight]{paper_files/figure-pdf/fig-scurve-true-sc-1.pdf}

\caption{\label{fig-scurve-true-sc}Two views of the true model (blue
lines) in \(2\text{-}D\) projections from \(7\text{-}D\), for the
S-curve data (black points). The data is spread along the S shape, and
does not vary much from this curve. (The \textbf{langevitour} software
is used to view the data with a tour, and the full video is available at
\url{https://youtu.be/I5GL23vLiw0}).}

@@

\includegraphics[width=1\textwidth,height=\textheight]{paper_files/figure-pdf/fig-NLDR-scurve-1.pdf}

@@

\caption{\label{fig-NLDR-scurve}Key steps for constructing the model on
the UMAP layout (\(k=2\)): (a) data, (b) hexagon bins, (c) bin
centroids, and (d) triangulated centroids. The S-curve data is shown.}

@@

Because we are working with distances between points, starting with data
having a standard scale, e.g.~{[}0, 1{]}, is recommended. The default
should take the aspect ratio produced by the NLDR
\((r_1, r_2, ..., r_k)\) into account. When \(k=2\), as in hexagon
binning, the default range is \([0, y_{i,\text{max}}], i=1,2\), where
\(y_{1,\text{max}}=1\) and \(y_{2,\text{max}} = \frac{r_2}{r_1}\)
(Figure~\ref{fig-NLDR-scurve}). If the NLDR aspect ratio is ignored then
set \(y_ {2,\text{max}} = 1\).

@@

By default, the bin centroid is used for describing a hexagon (as done
in Figure~\ref{fig-NLDR-scurve} (c)), but any measure of center, such as
a mean or weighted mean of the points within each hexagon, could be
used. The bin centers, and the binned data, are the two important
components needed to render the model representation in high dimensions.

@@

Delaunay triangulation \citep{lee1980, alb2024} is used to connect
points so that edges indicate neighbouring observations, in both the
NLDR layout (Figure~\ref{fig-NLDR-scurve} (d)) and the \pD{} model
representation. When the data has been binned the triangulation connectd
centroids. The edges preserve the neighborhood information when the
model is lifted into \pD{}.

@@

\begin{minipage}{0.25\linewidth}
\includegraphics{figures/scurve/umap_trimesh_layout.png}\end{minipage}%
%
\begin{minipage}{0.25\linewidth}
\includegraphics{figures/scurve/sc_umap_best_1.png}\end{minipage}%
%
\begin{minipage}{0.25\linewidth}
\includegraphics{figures/scurve/sc_umap_best_2.png}\end{minipage}%
%
\begin{minipage}{0.25\linewidth}
\includegraphics{figures/scurve/sc_umap_best_3.png}\end{minipage}%

\caption{\label{fig-scurve-sc}Model in \gD{}, on the UMAP layout, and
three views of the fit in projections from \(7\text{-}D\), for the
S-curve data (\((s_1, \ s_2) = (-0.160, \ -0.263)\),
\(b = 405 \  (15, \ 27)\), \(m = 70\), benchmark value to remove low
density hexagons is \(0.250\), and benchmark value to remove large edges
is \(0.159\)). MSE is \(0.0462\). The model closely fits the shape, but
it doesn't fully fill out the width of the S, which means that it does
not adequately capture the surface (The \textbf{langevitour} software is
used to view the data with a tour, and the full video is available at
\url{https://youtu.be/i7F7xpN1Hz8}).}

@@

\includegraphics[width=1\textwidth,height=0.3\textheight]{paper_files/figure-pdf/fig-p-d-error-in-2d-scurve-1.pdf}

\caption{\label{fig-p-d-error-in-2d-scurve}error}

@@

\includegraphics[width=1\textwidth,height=\textheight]{paper_files/figure-pdf/fig-predict-scurve-1.pdf}

\caption{\label{fig-predict-scurve}Comparison of prediction generated
using the exiting \texttt{umap} prediction method and our method: (a) A
view of the true model in projections from \(7\text{-}D\), (b) predicted
data from the \texttt{umap} prediction method, (c) predicted data from
our method, and (d) comparison of predictions from both methods. In plot
(d), points representing the same data but predicted by different
methods are connected by lines. Need to add concluded sentence after
changing the data.}

@@

The values \((s_1, \ s_2)\) define the position of the centroid of the
bottom left hexagon. By default, this is at \(s_1 = -q, s_2 = -qr_2\),
where \(q\) is the buffer sound the data. The choice of these values can
have some effect on the distribution of bin counts.
Figure~\ref{fig-param-scurve} (a) illustrates this. The distribution of
bin counts for \(s_1\) varying between \(-0.1-0.0\) is shown. Generally,
a more uniform distribution among these possibilities would indicate
that the bins are reliably capturing the underlying distribution of
observations.

\includegraphics[width=1\textwidth,height=\textheight]{paper_files/figure-pdf/fig-bins-scurve-1.pdf}

\caption{\label{fig-bins-scurve}Hexbin density plots of UMAP layout of
the S-curve data, using three different bin inputs: (a)
\(b = 91 \text{ } (7, \text{ }13)\), (b)
\(b = 220 \text{ } (11, \text{ }20)\), and (c)
\(b = 312 \text{ } (13, \text{ }24)\). Color indicates standardized
counts, dark indicating high count and light indicates low count. At the
smallest bin size the data segregates into two separate groups,
suggesting this is too many bins. Using the MSE of the model fit in
\(p\text{-}D\) helps decide on a useful choice of number of bins.}

The default number of bins \(b=b_1\times b_2\) is computed based on the
sample size, by setting \(b_1=n^{1/3}\), consistent with the
Diaconis-Freedman rule \citep{freedman1981}. The value of \(b_2\) is
determined analytically by \(b_1, q, r_2\). Values of \(b_1\) between
\(2\) and \(b_1 = \sqrt{\frac{n}{r_2}}\) are allowed.
Figure~\ref{fig-param-scurve} (b) shows the effect of different choices
of \(b_1\) on the MSE of the fitted model.

@@

By default, when assessing the choice of \(b_1\), the total number of
bins is measured by the number of \textbf{non-empty} bins. This more
accurately reflects the hexagon grid relative the MSE than the full
number of bins in the grid. It may also be beneficial to remove low
count bins also, in the situation where data is clustered or stringy,
where the observed data is sparse. In order to decide if this is
necessary, you would examine the distribution of bin counts, or the
density which puts the counts on a standard scale. If there is something
of a gap at low values, this would suggest a potential value to use as a
cutoff. Alternatively, one could choose to remove based on a percentile,
the bins with density in the lowest 5\% of all bins, for example.
Figure~\ref{fig-param-scurve} (c) illustrates the effect on the model
representation of removing bins below different percentages. Generally,
we would urge caution in removing low count bins.

The benchmark value for removing low-density hexagons ranges between
\(0\) and \(1\). When analyzing how these benchmark values influence
model performance, it's essential to observe the change in MSE as the
benchmark value increases (Figure~\ref{fig-param-scurve}). The MSE shows
a gradual decrease as the benchmark value goes from \(1\) to \(0\).
Evaluating this rate of increase is important. If the increment is not
considerable, the decision might lean towards retaining low-density
hexagons.

\includegraphics[width=1\textwidth,height=\textheight]{paper_files/figure-pdf/fig-param-scurve-1.pdf}

\caption{\label{fig-param-scurve}Various plots to help assess best
hexagon bin parameters, thresholds to remove low density bins and large
edges. Both (b) and (c) show MSE, against number of bins along the
x-axis and standardised count. A good benchmark value for these
parameters is when the MSE drops and then flattens out. Plot (a) shows
the distribution of stadardised counts of hexagons. Plot (d) shows the
distribution of \(2\text{-}D\) Euclidean distances between bin
centroids, with a good benchmark value for removing large edges would
being the distance that shows the first large decrease.}

@@

Edges define the neighbourhood structure, in order to provide a smooth
\gD{} representation of the fitted model.
Figure~\ref{fig-scurve-true-sc} shows a wire frame of the true model
that was used to generate the S-curve example data. The ideal is that
the representation of the fitted model, at least for this example where
we know the true model, should look similar to this.

@@

The decision on edge length removal is made based on the distribution of
edge lengths. In particular, a gap between values, where there a
concentration of small values and then a few larger values, likely
suggests a cutoff for edge removal. Because the triangulation is
typically done on the hexagon centroids, there are particular discrete
edge lengths, based on bin widths. Figure~\ref{fig-NLDR-scurve} (d)
illustrates edge length distributions.

@@

{\centering \includegraphics[width=1\textwidth,height=\textheight]{paper_files/figure-pdf/unnamed-chunk-41-1.pdf}

\caption{long edge removal}

@@

\caption{\label{fig-scurve-sc-best}Model in \gD{}, on the tSNE layout,
and three views of the fit in projections from \(7\text{-}D\), for the
S-curve data (\((s_1, \ s_2) = (-0.170, \ -0.137)\),
\(b = 256 \  (16, \ 16)\), \(m = 70\), benchmark value to remove low
density hexagons is \(0.250\), and benchmark value to remove large edges
is \(0.180\)). MSE is \(0.0432\). The model closely fits the shape, but
it has some twists (The \textbf{langevitour} software is used to view
the data with a tour, and the full video is available at
\url{https://youtu.be/ZuLdp89qJ6g}).}

>>>>

With our goal being to make a representation of this \gD{} layout that
can be lifted into high-dimensional space, the layout needs to be
augmented to include neighbour information. A simple approach would be
to triangulate the points and add edges. A more stable approach is to
first bin the data, reducing it from \(n\) to \(m\leq n\) observations,
and connect the bin centroids. We recommend using a hexagon grid because
it better reflects the data distribution and has less artifacts than a
rectangular grid. This process serves to reduce some noisiness in the
resulting surface shown in \pD{}. The steps in this process are shown in
Figure~\ref{fig-NLDR-two-curvy}, and documented below.

@@

@@

\includegraphics[width=1\textwidth,height=\textheight]{paper_files/figure-pdf/fig-NLDR-two-curvy-1.pdf}

@@

\caption{\label{fig-NLDR-two-curvy}Key steps for constructing the model
on the UMAP layout (\(k=2\)): (a) data, (b) hexagon bins, (c) bin
centroids, and (d) triangulated centroids. The S-curve data is shown.}

@@

Because we are working with distances between points, starting with data
having a standard scale, e.g.~{[}0, 1{]}, is recommended. The default
should take the aspect ratio produced by the NLDR
\((r_1, r_2, ..., r_k)\) into account. When \(k=2\), as in hexagon
binning, the default range is \([0, y_{i,\text{max}}], i=1,2\), where
\(y_{1,\text{max}}=1\) and \(y_{2,\text{max}} = \frac{r_2}{r_1}\)
(Figure~\ref{fig-NLDR-two-curvy}). If the NLDR aspect ratio is ignored
then set \(y_ {2,\text{max}} = 1\).

@@

By default, the bin centroid is used for describing a hexagon (as done
in Figure~\ref{fig-NLDR-two-curvy} (c)), but any measure of center, such
as a mean or weighted mean of the points within each hexagon, could be
used. The bin centers, and the binned data, are the two important
components needed to render the model representation in high dimensions.

@@

Delaunay triangulation \citep{lee1980, alb2024} is used to connect
points so that edges indicate neighbouring observations, in both the
NLDR layout (Figure~\ref{fig-NLDR-two-curvy} (d)) and the \pD{} model
representation. When the data has been binned the triangulation connectd
centroids. The edges preserve the neighborhood information when the
model is lifted into \pD{}.

@@

@@

@@

@@

The values \((s_1, \ s_2)\) define the position of the centroid of the
bottom left hexagon. By default, this is at \(s_1 = -q, s_2 = -qr_2\),
where \(q\) is the buffer sound the data. The choice of these values can
have some effect on the distribution of bin counts.
\textbf{?@fig-param-two\_curvy} (a) illustrates this. The distribution
of bin counts for \(s_1\) varying between \(-0.1-0.0\) is shown.
Generally, a more uniform distribution among these possibilities would
indicate that the bins are reliably capturing the underlying
distribution of observations.

The default number of bins \(b=b_1\times b_2\) is computed based on the
sample size, by setting \(b_1=n^{1/3}\), consistent with the
Diaconis-Freedman rule \citep{freedman1981}. The value of \(b_2\) is
determined analytically by \(b_1, q, r_2\). Values of \(b_1\) between
\(2\) and \(b_1 = \sqrt{\frac{n}{r_2}}\) are allowed.
\textbf{?@fig-param-two\_curvy} (b) shows the effect of different
choices of \(b_1\) on the MSE of the fitted model.

@@

By default, when assessing the choice of \(b_1\), the total number of
bins is measured by the number of \textbf{non-empty} bins. This more
accurately reflects the hexagon grid relative the MSE than the full
number of bins in the grid. It may also be beneficial to remove low
count bins also, in the situation where data is clustered or stringy,
where the observed data is sparse. In order to decide if this is
necessary, you would examine the distribution of bin counts, or the
density which puts the counts on a standard scale. If there is something
of a gap at low values, this would suggest a potential value to use as a
cutoff. Alternatively, one could choose to remove based on a percentile,
the bins with density in the lowest 5\% of all bins, for example.
\textbf{?@fig-param-two\_curvy} (c) illustrates the effect on the model
representation of removing bins below different percentages. Generally,
we would urge caution in removing low count bins.

The benchmark value for removing low-density hexagons ranges between
\(0\) and \(1\). When analyzing how these benchmark values influence
model performance, it's essential to observe the change in MSE as the
benchmark value increases (\textbf{?@fig-param-two\_curvy}). The MSE
shows a gradual decrease as the benchmark value goes from \(1\) to
\(0\). Evaluating this rate of increase is important. If the increment
is not considerable, the decision might lean towards retaining
low-density hexagons.

@@

Edges define the neighbourhood structure, in order to provide a smooth
\gD{} representation of the fitted model.
\textbf{?@fig-two\_curvy-true-sc} shows a wire frame of the true model
that was used to generate the S-curve example data. The ideal is that
the representation of the fitted model, at least for this example where
we know the true model, should look similar to this.

@@

The decision on edge length removal is made based on the distribution of
edge lengths. In particular, a gap between values, where there a
concentration of small values and then a few larger values, likely
suggests a cutoff for edge removal. Because the triangulation is
typically done on the hexagon centroids, there are particular discrete
edge lengths, based on bin widths. Figure~\ref{fig-NLDR-two-curvy} (d)
illustrates edge length distributions.

@@

@@

\caption{\label{fig-two_curvy-sc-best}Model in \gD{}, on the tSNE
layout, and three views of the fit in projections from \(7\text{-}D\),
for the S-curve data (\((s_1, \ s_2) = (-0.170, \ -0.137)\),
\(b = 256 \  (16, \ 16)\), \(m = 70\), benchmark value to remove low
density hexagons is \(0.250\), and benchmark value to remove large edges
is \(0.180\)). MSE is \(0.0432\). The model closely fits the shape, but
it has some twists (The \textbf{langevitour} software is used to view
the data with a tour, and the full video is available at
\url{https://youtu.be/ZuLdp89qJ6g}).}
